---
title: "Introduction to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*R* is an interactive programming language emphasizing statistical
analysis and visualization. It is very useful for both light-weight
data manipulation, and for advanced statistical analysis and
visualization. *R* is available as Free Software under the GNU General
Public License. It is highly extensible through user-contributed
'packages', of which there are more than 20,000 available through
[CRAN][] (Comprehensive *R* Archive Network) and more than 2,000
through [Bioconductor][] (for high-throughput genomic data).

[CRAN]: https://CRAN.R-project.org
[Bioconductor]: https://bioconductor.org

# *RStudio*

## Panes

## Scripts

## Markdown

# *R*

## Vectors

## `data.frame`

## Functions

## Packages

Examples:

- [dplyr][]: data manipulation based on 'tidy' principles.
- [ggplot2][]: plotting using the 'grammar of graphics'

Installation

Use

```{r, message = FALSE}
library(dplyr)
```

`tibble` versus `data.frame`

Pipes

[dplyr]: https://CRAN.R-project.org/package=dplyr
[ggplot2]: https://CRAN.R-project.org/package=dplyr

## Help!!!

*RStudio*

*R*

- For help on individual commands, use `?` at the command line, e.g.,
  `?data.frame`
- For package help, try

    ```{r, eval = FALSE}
    vignette(package = "dplyr")
    #> Vignettes in package 'dplyr':
    #> 
    #> colwise                 Column-wise operations (source, html)
    #> base                    dplyr <-> base R (source, html)
    #> grouping                Grouped data (source, html)
    #> dplyr                   Introduction to dplyr (source, html)
    #> programming             Programming with dplyr (source, html)
    #> rowwise                 Row-wise operations (source, html)
    #> two-table               Two-table verbs (source, html)
    #> in-packages             Using dplyr in packages (source, html)
    #> window-functions        Window functions (source, html)
    #> 
    vignette(package = "dplyr", "dplyr")
    ```

Other sources of help

- Google
- [StackOverflow][]
- ChatGPT & other AI

[StackOverflow]: https://stackoverflow.com/questions/tagged/r

# Session information

```{r}
sessionInfo()
```
