---
title: "Introduction to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*R* is an interactive programming language emphasizing statistical
analysis and visualization. It is very useful for both light-weight
data manipulation, and for advanced statistical analysis and
visualization. *R* is available as Free Software under the GNU General
Public License. It is highly extensible through user-contributed
'packages', of which there are more than 20,000 available through
[CRAN][] (Comprehensive *R* Archive Network) and more than 2,000
through [Bioconductor][] (for high-throughput genomic data).

[CRAN]: https://CRAN.R-project.org
[Bioconductor]: https://bioconductor.org

# *RStudio*

## Panes

## Scripts

## Markdown

# *R*

```{r}
1 + 2
```

```{r}
sqrt(2)
```

## Vectors and variables

*R* usually works on *vectors* of values

```{r}
c(1, 2, 3)
```

It usually makes sense to assign a vector of values to a variable, to
make it easy to repeatedly reference the same set of values

```{r}
x <- c(1, 2, 3)
x
```

Types of vectors

- logical -- `c(TRUE, FALSE)`
- integer -- `c(1, 2, 3)`
- numeric -- `c(1.41, 2.30, 3.14159)`
- character -- `c("May", "June", "July")`
- complex

Statistical concepts

- 'Missing' values
  - use `NA` in any vector, e.g., `c(1, 2, NA, 3)`
- factor with levels
  - Often, in an experimental design, there are defined number of
    groups, e.g., 'treatment' and 'control'
  - Use a `factor` to represent these values
  
      ```{r}
      levels <- c("treatment", "control") # possible values
      ## observed values -- four samples
      values <- c("treatment", "treatment", "control", "control")
      factor(values, levels = levels)
      ```

Operations, e.g., numeric

- two vectors -- element-wise

    ```{r}
    x <- c(1, 2, 3)
    y <- c(4, 5, 6)
    x + y
    ```

- vector and scalar -- scalar is 'recycled' to the same length as
  vector, and then element-wise operations are preformed
  
  ```{r}
  x * 2
  ```

Common operations

- logical -- `&` ('and'), `|` ('or'), `!` (not), `==` (element-wise equivalence)

    ```{r}
    c(TRUE, TRUE, FALSE, FALSE) & c(TRUE, FALSE, TRUE, FALSE)
    c(TRUE, TRUE, FALSE, FALSE) | c(TRUE, FALSE, TRUE, FALSE)
    !c(TRUE, TRUE, FALSE, FALSE)
    c(TRUE, TRUE, FALSE, FALSE) == c(TRUE, FALSE, TRUE, FALSE)
    ## Fun!
    c(TRUE, FALSE, NA) & c(NA, NA, NA)
    c(TRUE, FALSE, NA) | c(NA, NA, NA)
    ```

- integer / numeric -- `+`, `-`, `*`. Comparison `==`, `>`, `>=`, `<`,
  `<=`. Many functions like `log()`, `sqrt()`, ...

- character -- `paste()` (paste two string togther), `nchar()`, `substring()`

    ```{r}
    month <- c("April", "May", "June", "July")
    year <- c("2020", "2021", "2022", "2023")
    paste(year, month)
    ```

Subsetting

```{r}
x <- c(1, 2, 3, 4)
x[c(1, 3)]
x[x >= 3]
```

## `data.frame`

Organizing vectors into columns

```{r}
df <- data.frame(
    Month = month, Year = year,
    Temperature = c(20, 22, 24, 25)
)
df
```

Subsetting

- columns 

    ```{r}
    df[, c("Month", "Temperature")]
    ```

- rows

    ```{r}
    df[c(1, 3), ]
    ```

Column access

- `$`; can be useful to subset

    ```{r}
    df$Temperature
    df[df$Temperature > 22,]
    ```

- `[[`; useful when a variable defines the column of interest

    ```{r}
    column_of_interest <- "Temperature"
    df[[ column_of_interest ]]
    ```

## Functions

```{r}
x <- rnorm(100)
x
```

Look up the help page for `rnorm` -- `?rnorm`

Plot a histogram using `hist()`

```{r}
x <- rnorm(1000)
hist(x)
```

Generate two equal-length vectors and visualize as a scatter plot

```{r, fig.asp = 1}
y <- x + rnorm(1000) # element-wise addition; each element of 'y' is
                     # the sum of two random variables
plot(x, y)           # one way to plot
plot(y ~ x)          # another way to plot -- 'y as a function of x'
```

Place `x` and `y` into a `data.frame`, and use to plot

```{r, fig.asp = 1}
df <- data.frame(X = x, Y = y)
plot(Y ~ X, df)
```

Subset to just the positive quadrant

```{r, fig.asp = 1}
plot(Y ~ X, df[df$X > 0 & df$Y > 0,])
```

A built-in data frame -- `mtcars`

```{r}
mtcars
```

Plot miles per gallon `mpg` as a function of number of cylinders, `cyl`

```{r, fig.asp = 1}
plot(mpg ~ cyl, mtcars) # scatter plot
```

Note though that `cyl` should probably be a factor -- a finite number
of possible values.

```{r}
mtcars$cyl
factor(mtcars$cyl) # 'levels' have a sensible default
```

Treat `cyl` as a factor when plotting

```{r, fig.asp = 1}
plot(mpg ~ factor(cyl), mtcars)
```

## Packages

Examples:

- [dplyr][]: data manipulation based on 'tidy' principles.
- [ggplot2][]: plotting using the 'grammar of graphics'

Installation

- Packages only need to be installed once

    ```{r, eval = FALSE}
    ## also possible to use RStudio 'Packages' panel...
    install.packages("dplyr", repos = "https://CRAN.R-project.org")
    ```

Use

- Attach packages in each *R* session where you would like to use it

    ```{r, message = FALSE}
    library(dplyr)
    ```

`tibble` versus `data.frame`

- only some values displayed -- really useful even for this small data
  set
- column data types indicated

    ```{r}
    tbl <- tibble(mtcars)
    tbl
    ```

Pipes

- `|>` takes the value on the 'left-hand side' and uses it as the
  first argument in the function on the 'right-hand side'
  
  ```{r}
  tbl <-
      mtcars |>
      tibble()
  ```

- this can be a very useful paradigm -- 'take the mtcars data set, and
  apply the tibble function to it' -- and allows a series of related
  operations to be chained together

[dplyr]: https://CRAN.R-project.org/package=dplyr
[ggplot2]: https://CRAN.R-project.org/package=dplyr

## Help!!!

*RStudio*

*R*

- For help on individual commands, use `?` at the command line, e.g.,
  `?data.frame`
- For package help, try

    ```{r, eval = FALSE}
    vignette(package = "dplyr")
    #> Vignettes in package 'dplyr':
    #> 
    #> colwise                 Column-wise operations (source, html)
    #> base                    dplyr <-> base R (source, html)
    #> grouping                Grouped data (source, html)
    #> dplyr                   Introduction to dplyr (source, html)
    #> programming             Programming with dplyr (source, html)
    #> rowwise                 Row-wise operations (source, html)
    #> two-table               Two-table verbs (source, html)
    #> in-packages             Using dplyr in packages (source, html)
    #> window-functions        Window functions (source, html)
    #> 
    vignette(package = "dplyr", "dplyr")
    ```

Other sources of help

- Google
- [StackOverflow][]
- ChatGPT & other AI

    - can be amazingly helpful, especially if asked to perform specific tasks
    - also useful to 'explain this code' or 'add comments to this code'
    - what out for halucinations!

[StackOverflow]: https://stackoverflow.com/questions/tagged/r

# Session information

```{r}
sessionInfo()
```
