---
title: "Data input and manipulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data input and manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Data input

## From Excel to 'CSV'

## CSV to R

## Example data: BRFSS

Download the data we will work with

```{r, eval = FALSE}
url <- "https://raw.githubusercontent.com/mtmorgan/BERD/main/inst/extdata/BRFSS-subset.csv"
destination <- file.path("data", basename(url))
if (!file.exists(destination)) {
    dir.create("data")
    download.file(url, destination)
}
```

```{r, echo = FALSE}
destination <- system.file(package = "BERD", "extdata", "BRFSS-subset.csv")
```

Read the data into *R* as a tibble using the [readr][] package.

```{r}
brfss <- readr::read_csv(destination)
brfss
```

# Explore the data using [dplyr][]

Attach [dplyr][] to our current *R* session.

```{r, message = FALSE}
library(dplyr)
```

## `count()`

```{r}
brfss |>
    count(Year)
brfss |>
    count(Sex)
brfss |>
    count(Year, Sex)
```

## `summarize()`

```{r}
brfss |>
    summarize(
        n = n(),
        ave_age = mean(Age, na.rm = TRUE),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_ht = mean(Height, na.rm = TRUE)
    )
```

## `group_by()`

```{r}
brfss |>
    group_by(Year, Sex) |>
    summarize(
        n = n(),
        ave_age = mean(Age, na.rm = TRUE),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_ht = mean(Height, na.rm = TRUE)
    )
```    

## `mutate()`

`Year` and `Sex` should really be factors. Also, add a column
for log-10 transformed weight.

```{r}
brfss_clean <-
    brfss |>
    mutate(
        Year = factor(Year, levels = c("1990", "2010")),
        Sex = factor(Sex, levels = c("Female", "Male")),
        Log10Weight = log10(Weight)
    )
brfss_clean
```

## `filter()`

Create a subset of the data that includes only Male samples, or only
Female samples from 2010.

```{r}
brfss_male <-
    brfss_clean |>
    filter(Sex == "Male")
brfss_male
brfss_female_2010 <-
    brfss_clean |>
    filter(Sex == "Female", Year == "2010")
brfss_female_2010
```

## `select()`

Use `select()` to choose particular columns

```{r}
brfss_male |>
    select(Year, Age, Weight, Log10Weight, Height)
```

# Initial visualization

What is the relationship between Height and Weight in Female samples
from 2010?

```{r, fig.asp = 1}
plot(Weight ~ Height, brfss_female_2010)
```

With Weight on a log scale?

```{r, fig.asp = 1}
plot(Weight ~ Height, brfss_female_2010, log = "y")
## another way, but maybe the Y-axis scale is harder to interpret?
## plot(Log10Weight ~ Height, brfss_female_2010)
```

Do Males differ in weight between 1990 and 2010, plotted on a log scale?

```{r, fig.asp = 1}
brfss_male |>
    group_by(Year) |>
    summarize(
        n = n(),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_log10_wt = mean(Log10Weight, na.rm = TRUE)
    )
plot(Weight ~ Year, brfss_male, log = "y")
```

# Session information

```{r}
sessionInfo()
```

[dplyr]: https://CRAN.R-project.org/package=dplyr
[readr]: https://CRAN.R-project.org/package=readr
