---
title: "Statistical Analysis and Additional Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Analysis and Additional Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Recap: BRFSS data input

Download (if necessary) and input the data from day 2 using [readr][]

```{r, eval = FALSE}
url <- "https://raw.githubusercontent.com/mtmorgan/BERD/main/inst/extdata/BRFSS-subset.csv"
destination <- file.path("data", basename(url))
if (!file.exists(destination)) {
    dir.create("data")
    download.file(url, destination)
}
```

```{r, echo = FALSE}
destination <- system.file(package = "BERD", "extdata", "BRFSS-subset.csv")
```

```{r, message = FALSE}
brfss <- readr::read_csv(destination)
```

Attach the [dplyr][] package and perform the same initial data
transformations, and create subsets of data.

```{r, message = FALSE}
library(dplyr)

brfss_clean <-
    brfss |>
    mutate(
        Year = factor(Year, levels = c("1990", "2010")),
        Sex = factor(Sex, levels = c("Female", "Male")),
        Log10Weight = log10(Weight)
    )

brfss_male <-
    brfss_clean |>
    filter(Sex == "Male")

brfss_female_2010 <-
    brfss_clean |>
    filter(Sex == "Female", Year == "2010")
```

# Statistical analysis

## t-test

We saw that Males in 2010 seemed to be heavier than Males in 1990

```{r, fig.asp = 1}
brfss_male |>
    group_by(Year) |>
    summarize(
        n = n(),
        ave_wt = mean(Weight, na.rm = TRUE),
        ave_log10_wt = mean(Log10Weight, na.rm = TRUE)
    )
plot(Weight ~ Year, brfss_male, log = "y")
```

But is this difference statistically significant? Answer this using a
*t*-test to compare the mean of two groups.

Start by consulting the help page `?t.test` with 'Usage' labeled "S3
method for class 'formula'". Use the `Log10Weight` column so that we
are comparing log-transformed Weight.

```{r}
t.test(Log10Weight ~ Year, brfss_male)
```

The *p*-value displayed here is `< 2.2e-16`, so we are *very*
confident that Males in 2010 are heavier than Males in 1990.

The 'sample estimates' agree with the values we calculated using
`group_by()` and `summarize()`; this provides us with some confidence
that we have not done something very wrong.

Note that `t.test()` actually returns an *object* that we can capture
and manipulate, e.g., to obtain a vector of group mean estimates.

```{r}
t_test <- t.test(Log10Weight ~ Year, brfss_male)
t_test$estimate
```

## Linear regression

It seems like there was a positive relationship between Height and
(log 10) Weight in Females in 2010.

```{r, fig.asp = 1}
plot(Weight ~ Height, brfss_female_2010, log = "y")
```

Fitting a linear regression in *R* is a two-step process

- Use `lm()` to fit the linear model
- Use `summary()` or `anova()` to assess the statistical significance
  of the fitted model.
  
Start by fitting the model, using notation like that used to plot the
data; use `Log10Weight` so the fit is to the log-transformed data.

```{r}
fit <- lm(Log10Weight ~ Height, brfss_female_2010)
fit
```

The `fit` object doesn't initially seem very useful -- it tells us
what we did, and summarizes the intercept (`r fit$coefficients[[1]]`)
and slope (`r fit$coefficients[[2]]`), but does not provide an ANOVA
table or other summary about whether the fit is statistically
significant.

Use `anova()` to obtain an ANOVA table summarizing the fit.

```{r}
anova(fit)
```

The table indicates that there is a highly significant relationship
(`Pr(>F)` is `< 2.2e-16`) between Weight and Height. `summary()`
provides the ANOVA table and additional information

```{r}
summary(fit)
```

e.g., showing that about 10% ('Adjusted R-squared') of the variation
in log 10 Weight is explained by Height. There are likely many other
factors contributing to Weight variation.

To visualize the linear regression, plot the data points and then use
`abline()` to add the regression line; we also add a 'main' title to
the plot.

```{r, fig.asp = 1}
plot(
    Log10Weight ~ Height, brfss_female_2010,
    main = "Log 10 Weight vs Height, BRFSS Female 2010"
)
abline(fit, col = "red", lwd = 4) # red color, 4x default line width
```

# Additional analysis and visualization

Let's use the [ggplot2][] package to perform additional analysis and
visualization

```{r, message = FALSE}
library(ggplot2)
```

[ggplot2][] implements a 'grammar of graphics'. The author of the
package has provided a unique way to create plots -- start with a
`ggplot()` graph, and then add `aes()`thetics (what will be plotted)
and `geom()`etries (how to plot the aesthetics), and additional layers
to, e.g., transform the y-axis to a log scale.

```{r, fig.asp = 1, warning = FALSE}
ggplot(brfss_female_2010) +
    aes(x = Height, y = Weight) +
    geom_point() +
    scale_y_log10()
```

Add a linear regression using `geom_smooth(method = "lm")`

```{r, fig.asp = 1, warning = FALSE}
ggplot(brfss_female_2010) +
    aes(x = Height, y = Weight) +
    geom_point() +
    geom_smooth(method = "lm") +
    scale_y_log10()
```

Note that the fitted regression includes confidence bands, which can
be very helpful when trying to interpret the fit.

Let's not specify a 'method' for `geom_smooth()`, letting [ggplot2][]
use it's default method...

```{r, fig.asp = 1, warning = FALSE}
ggplot(brfss_female_2010) +
    aes(x = Height, y = Weight) +
    geom_point() +
    geom_smooth() +
    scale_y_log10()
```

[ggplot2][] has chosen to use a 'generalized additive model' that fits
a smoothed curve. This makes fewer *a priori* assumptions about the
relationship between Weight and Height, and in particular suggests
that (a) the linear relationship is appropriate for the central part
of the data, but not at the extremes of Height and (b) there is
considerable uncertainty at the extremes of Height. This approach
(making fewer assumptions about the relationship between variables) is
very appropriate in early stages of exploratory analysis, where one is
still trying to understand data without making additional or
unnecessary assumptions.

# Using AI

# Session information

```{r}
sessionInfo()
```

[dplyr]: https://CRAN.R-project.org/package=dplyr
[readr]: https://CRAN.R-project.org/package=readr
[ggplot2]: https://CRAN.R-project.org/package=ggplot2
